\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
working on LMEM vs GAM part
Working on confidence interval, need to center the comparison to zero

@@

```{r, message = FALSE,include=FALSE}
library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(ggsci)
library(mgcv)

@@

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(l-q-response-caption)' }
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

@@

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.

@@

One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?". The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.

@@

## plot the mean response
f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
    geom_line(size=1,show.legend = FALSE)+
    geom_point(show.legend = FALSE)+
  labs(y=expression(paste('Mean StO'[2],'\n (real)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
  plot_layout(tag_level = 'new')

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
    geom_point(show.legend=FALSE)+
    stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",size=1,show.legend = FALSE)+
  labs(y=expression(paste('StO'[2],'\n (simulated)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

@@

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)

@@

#lm2<-lme(StO2_sim ~ Group + Day + Group:Day,
#                 data = dat_sim,
#                 random = ~ 1 | subject,
#                 correlation = corCompSymm(form = ~ 1 | subject)
#  )

#creates a dataframe using the length of the covariates for the GAM
gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#creates a dataframe using the length of the covariates for the rm-ANOVA
lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = c(0,2,5,7,10),
                        subject=factor(rep(1:10)),
                          )
lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=gam_predict,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
      scale_color_aaas()+
  scale_fill_aaas()

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=lm_predict,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
      scale_color_aaas()+
  scale_fill_aaas()

sim_smooth_plot<-f3 & theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )

sim_linear_plot<-f4 & theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )

@@

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment. B: Simulated data from  a normal distribution with standard deviation of 10% with 10 observations per time point. In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95% confidence intervals.

f2+inset_element(f1, left = 0.01, bottom = 0.7, right = 0.5, top = 1.0)+f3+f4+
  plot_layout(guides = 'collect')+
   plot_annotation(tag_levels='A')&
   ylim(c(0,75)) &
   theme(
      text=element_text(size=18),
      legend.title=element_blank(),
      legend.text = element_blank()
      )&
  scale_color_aaas()

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data that has missing observations. Again, using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and a GAM model with interaction can be fitted to the remaining observations.

@@

#plot the simulated data
f4<-ggplot(dat_missing, aes(x = Day, y = StO2_sim, color = Group)) +
    geom_point() +
  stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",size=1)+
    ggtitle("Missing data") +
  labs(y='StO2')+
  scale_color_aaas()+
    theme_classic()

f2+f4+plot_layout(guides = 'collect') &
  ylim(c(0,75)) &
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )&
  scale_color_aaas()

#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
#appraise the model
appraise(mod_m1)

@@

ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    facet_wrap(~Group)+
    geom_point(colour='black',size=1,alpha=0.5)+
    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
                data=m_predict, alpha=0.5,
                inherit.aes=FALSE) +
    geom_line(aes(y=(fit),color=factor(Group)),
              size=1,data=m_predict,show.legend = FALSE)

comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', 
         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
                              )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme_classic()+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )
```

# pdat <- expand.grid(Day = seq(0, 10, length = 400),
#                     Group = c('Treatment', 'Control'))
#
# smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
#                         unconditional = FALSE) {
#     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
#     c1 <- grepl(f1, colnames(xp))
#     c2 <- grepl(f2, colnames(xp))
#     #r1 <- newdata[[var]] == f1
#     #r2 <- newdata[[var]] == f2
#     r1 <- with(newdata, Group == f1)
#     r2 <- with(newdata, Group == f2)
#     ## difference rows of xp for data from comparison
#     X <- xp[r1, ] - xp[r2, ]
#     ## zero out cols of X related to splines for other lochs
#     X[, ! (c1 | c2)] <- 0
#     ## zero out the parametric cols
#     X[, !grepl('^s\\(', colnames(xp))] <- 0
#     dif <- X %*% coef(model)
#     se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
#     crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
#     upr <- dif + (crit * se)
#     lwr <- dif - (crit * se)
#     data.frame(pair = paste(f1, f2, sep = '-'),
#                diff = dif,
#                se = se,
#                upper = upr,
#                lower = lwr)
# }
#
# comp1<-smooth_diff(mod_int2,pdat,'Control','Treatment')
#
# comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
#                    rbind(comp1))
#
# c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
#     geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
#     geom_line(color='black',size=1) +
#     facet_wrap(~ pair) +
#   theme_classic()+
#     labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
#   scale_x_continuous(breaks=c(0,2,5,7,10))+
#   theme(
#      text=element_text(size=18),
#      legend.title=element_blank()
#      )
#
#
#
# c1

@@

>>>>

# Abstract

In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA). However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations. When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference. In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R. This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R. The code and data used to generate this paper are available at:

@@

```{r, message = FALSE,include=FALSE}
library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(ggsci)
library(mgcv)
library(gratia)

@@

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

@@

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.

@@

One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?". The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C. This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .

@@

## plot the mean response
f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
    geom_line(size=1,show.legend = FALSE)+
    geom_point(show.legend = FALSE)+
  labs(y=expression(paste(StO[2],' (real)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,5,10))+
    scale_y_continuous(breaks=c(0,40))+
  plot_layout(tag_level = 'new')+
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.text=element_text(size=14)
  )

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
    geom_point(show.legend=FALSE)+
    stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",size=1,show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

@@

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

@@

#creates a dataframe using the length of the covariates for the GAM
gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1),
                         subject=factor(rep(1:10)))

#creates a dataframe using the length of the covariates for rm-ANOVA
lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = c(0:10),
                        subject=factor(rep(1:10)),
                          )
lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=gam_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

@@

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.

@@

#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
#appraise the model
appraise(mod_m1)

@@

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment. Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point. Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment. Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time. D: Smooths for the GAM model for the simulated data with 40% of its observations missing. Lines represent smooths for each group, shaded regions are 95% confidence intervals

mult_plot<-f2+inset_element(
  f1, left = 0.01, 
  bottom = 0.5, 
  right = 0.5, 
  top = 1.0)+
  f3+f4+f6+
  #plot_layout(guides = 'collect')+
   plot_annotation(tag_levels='A')&
   ylim(c(-5,75)) &
  theme(
     text=element_text(size=18)
     )&
  scale_color_aaas()

mult_plot

@@

pdat <- expand.grid(Day = seq(0, 10, length = 400),
                    Group = c('Treatment', 'Control'))

smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    #r1 <- newdata[[var]] == f1
    #r2 <- newdata[[var]] == f2
    r1 <- with(newdata, Group == f1)
    r2 <- with(newdata, Group == f2)
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

comp1<-smooth_diff(gam1,pdat,'Control','Treatment')

comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp1)) %>%
  mutate(interval=case_when(
    upper>0 & lower<0~"no-diff",
    upper<0~"less",
    lower>0~"greater"
  ))

c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
  theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )
#+
 # geom_rect(aes(color=interval))

c1

###for missing data
comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', 
         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
                              )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme_classic()+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Working on confidence interval, need to center the comparison to zero

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{background}{%
\section{Background}\label{background}}

@@

\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
\end{figure}

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for time and original datapoints. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
\end{figure}

@@

\hypertarget{gams-and-covariance}{%
\subsection{GAMs and covariance}\label{gams-and-covariance}}

Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}52{]}. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time. An example where this is covered is in the Appendix.

@@

>>>>

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA). However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations. When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference. In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the \emph{mgcv} package in R. This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R. The code and data used to generate this paper are available at:

\hypertarget{background}{%
\section{Background}\label{background}}

@@

\caption{(l-q-response-caption)}\label{fig:l-q-response}
\end{figure}

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
\end{figure}

@@

\hypertarget{the-analyisis-of-longitudinal-biomedical-data-using-gams}{%
\section{The analyisis of longitudinal biomedical data using GAMs}\label{the-analyisis-of-longitudinal-biomedical-data-using-gams}}

The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \texttt{mgcv}. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

\hypertarget{simulated-data}{%
\subsection{Simulated data}\label{simulated-data}}

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}vishwanath2009{]}. An increase in oxygen saturation in the ``Treatment'' group (chemotherapy) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.

\hypertarget{an-interaction-gam-model-for-longitudinal-data}{%
\subsection{An interaction GAM model for longitudinal data}\label{an-interaction-gam-model-for-longitudinal-data}}

An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time. Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}52{]}.

For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in oxygen saturation. The model then needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( \emph{Day} ) and that the other is a factor ( \emph{Group} ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using \(\textsf{R}\) and the package \texttt{mgcv} the model syntax is:

\texttt{m1\textless{}-gam(StO2\_sim\textasciitilde{}Group+s(Day,by=Group,k=5),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}

This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation \texttt{StO2\_sim} is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots. The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(StO_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B). Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \texttt{gratia}{[}54{]}. A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}37{]} and {[}55{]}.

One question might be ``what is the fit that an rm-ANOVA model produces for the simulated data?''. The rm-ANOVA model, which corresponds to \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C. This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .

Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists. Using the simulated \(StO_2\) values from Figure (\ref{fig:sim-smooth-plot}, B), 40\% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.

{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/sim-smooth-plot-1}

\caption{Simulated data and smooths for oxygen saturation in tumors. A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment. B: Simulated data from a normal distribution with standard deviation of 10\% with 10 observations per time point. In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
\end{figure}

@@

\leavevmode\hypertarget{ref-gratia}{}%
{[}54{]} G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.

\leavevmode\hypertarget{ref-harezlak2018}{}%
{[}55{]} J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.
